---
output:
  html_document: default
  pdf_document: default
---
EXPLORATORY DATA ANALYSIS by Kemal Kar
========================================================



```{r echo=FALSE, message=FALSE, warning=FALSE, packages}

#Loading libraries needed to run the functions.

library(ggplot2)
library(gridExtra)
library(GGally)
library(dplyr)
library(memisc)

```

```{r echo=FALSE, Load_the_Data}

# Loading Data

RedWineQuality <- read.csv('wineQualityReds.csv')
head(RedWineQuality)

```



```{r echo=FALSE}

# Get the names of the columns

names(RedWineQuality)

```

# Univariate Plots Section


```{r}

summary(RedWineQuality)

```



```{r echo=FALSE, Univariate_Plots}

#Plot the histogram for quality 

qplot(x=quality, data=RedWineQuality, binwidth = 1) +
  scale_x_continuous(breaks=seq(2,8,1)) +
  theme(text=element_text(size=10), plot.title=
          element_text(size=12, face='bold',hjust = 0.5)) +
  labs(title='Quality Histogram', x= 'Quality', y = 'Count')

```

```{r, message=FALSE, echo=FALSE, warning=FALSE }

qplot(x=pH, data=RedWineQuality, binwidth = 0.02) +
  scale_x_continuous(limits=c(2.9,3.7)) +
  theme(text=element_text(size=10), plot.title=
          element_text(size=12, face='bold',hjust = 0.5)) +
  labs(title='pH Histogram', x= 'pH', y = 'Count')

```

```{r, echo=FALSE}
 
qplot(x=fixed.acidity, data=RedWineQuality, binwidth = 0.2) +
  scale_x_continuous(limits=c(4,14)) +
  theme(text=element_text(size=10), 
        plot.title=element_text(size=12, face='bold',hjust = 0.5)) +
  labs(title='Fixed Acidity Histogram', x= 'Fixed Acidity', y = 'Count')

```

```{r, echo=FALSE}

qplot(x=volatile.acidity, data=RedWineQuality, binwidth = 0.02) +
  scale_x_continuous(limits=c(0.1,1)) +
  theme(text=element_text(size=10), 
        plot.title=element_text(size=12, face='bold',hjust = 0.5)) +
  labs(title='Volatile Acidity Histogram', x= 'Volatile Acidity', y = 'Count')

```

```{r, echo=FALSE}

summary(RedWineQuality$citric.acid)

summary(log10(RedWineQuality$citric.acid+1))

summary(sqrt(RedWineQuality$citric.acid))

p_c1 <- qplot(x=citric.acid, data=subset(RedWineQuality, 
                                         !is.na(citric.acid)), binwidth = 0.01) +
  scale_x_continuous(limits=c(0,0.75)) +
  scale_y_continuous(limits=c(0,75)) +
  theme(text=element_text(size=10), 
        plot.title=element_text(size=12, face='bold',hjust = 0.5)) +
  labs(title='Citric Acid Histogram', x= 'Citric Acid', y = 'Count')


p_c2 <- qplot(x=log10(RedWineQuality$citric.acid+1),
              data=RedWineQuality, binwidth = 0.01) +
  scale_x_continuous(limits=c(0,0.3)) +
  theme(text=element_text(size=10), plot.title=
          element_text(size=12, face='bold',hjust = 0.5)) +
  labs(title='Log of Citric Acid Histogram', x= 'log(Citric Acid)', y = 'Count')


p_c3 <- qplot(x=sqrt(RedWineQuality$citric.acid), 
              data=RedWineQuality, binwidth = 0.02) +
  scale_x_continuous(limits=c(0.2,0.8)) +
  theme(text=element_text(size=10), plot.title=
          element_text(size=12, face='bold',hjust = 0.5)) +
  labs(title='Sqrt of Citric Acid Histogram', x='Sqrt(Citric Acid)', y='Count')

grid.arrange(p_c1,p_c2,p_c3,ncol=2)

```

```{r, echo=FALSE}

qplot(x=residual.sugar, data=RedWineQuality, binwidth = 0.1) +
  scale_x_continuous(limits=c(0.9,4)) +
  theme(text=element_text(size=10), plot.title=
          element_text(size=12, face='bold',hjust = 0.5)) +
  labs(title='Residual Sugar Histogram', x= 'Residual Sugar', y = 'Count')
   
```


```{r, echo=FALSE}

qplot(x=chlorides, data=RedWineQuality, binwidth = 0.003) +
  scale_x_continuous(limits=c(0.02,0.12)) +
  theme(text=element_text(size=10), plot.title=
          element_text(size=12, face='bold',hjust = 0.5)) +
  labs(title='Chlorides Histogram', x= 'Chlorides', y = 'Count')

```

```{r, echo=FALSE}


qplot(x=free.sulfur.dioxide, data=RedWineQuality, binwidth = 1) +
  scale_x_continuous(limits=c(1,45)) +
  theme(text=element_text(size=10), plot.title=
          element_text(size=12, face='bold',hjust = 0.5)) +
  labs(title='Free Sulfur Dioxide Histogram', x='Free Sulfur Dioxide', y='Count')


```

```{r, echo=FALSE}

qplot(x=total.sulfur.dioxide, data=RedWineQuality, , binwidth = 3) +
  scale_x_continuous(limits=c(6,150)) +
  theme(text=element_text(size=10), plot.title=
          element_text(size=12, face='bold',hjust = 0.5)) +
  labs(title='Total Sulfur Dioxide Histogram',
       x='Total Sulfur Dioxide', y='Count')

```


```{r, echo=FALSE}

qplot(x=alcohol, data=RedWineQuality, binwidth = 0.2) +
  scale_x_continuous(limits=c(8.4,14)) +
  theme(text=element_text(size=10), plot.title=
          element_text(size=12, face='bold',hjust = 0.5)) +
  labs(title='Alcohol Histogram', x= 'Alcohol', y = 'Count')

```

# Univariate Analysis

> **Tip**: Now that you've completed your univariate explorations, it's time to
reflect on and summarize what you've found. Use the questions below to help you
gather your observations and add your own if you have other thoughts!

### What is the structure of your dataset?

Quality is the response variable in the dataset. The histogram displays that 
it is in normal distribution.

The explanatory variables in this dataset are quantitative. Some of them are 
explored in univariate plots:

1. pH, fixed and volatile acidity, residual sugar and chlorides values are also
in normal distribution, similar to quality values.

2. Alcohol, total and free sulfur dioxide histograms are right skewed. 

### What is/are the main feature(s) of interest in your dataset?

Quality of the wine is a result of sulfur, acidity, sugar and chloride variables.
Throughout the analysis below, I will assess the correlation between the quality
and these variables.

### What other features in the dataset do you think will help support your \

Quality and the 1st group of variables above are in normal distribution. I will 
test whether this similarity in distribution means there is  strong correlation 
between these variables and the resultant quality.

pH value can be a resultant value of fixed, volalite acidity and citric acid 
variables. I will assess whether there is a relationship between these variables and pH value.

Total and free sulfur dioxide distibutions are very similarly right skewed. 
I will also check whether these 2 variables are strongly corelated. 

### Did you create any new variables from existing variables in the dataset?

At first check, no new variables are created.

### Of the features you investigated, were there any unusual distributions? \

Throughout the histograms, I removed the long tails of some of the variables by
cutting from the 1st and 3rd quadrants so that I could focus on the main 
distribution omitting the outliers.

Citric acid values seem to be distributed throughout the histogram plot. 
I checked the log and sqrt distributions if there could be a gaussian 
distribution hidden. 


# Bivariate Plots Section

In this section, The correlation between different explanatory variables and the
resultant variable, quality.

Let's start with creating a heatmap which displays the correlations between 
different variables. 

```{r, echo=FALSE}

ggcorr(data=RedWineQuality) +  theme(text=element_text(size=8))

```

In the heatmap, the following correlations can be observed:
1. In higher quality wines, the alcohol , sulphates and citric acid is found more.
There is a positive relationship between quality and these variables.
2. Volatile acidity is found less in hihger quality wines where as citric acid 
increases. 
3. pH value of the wine is negatively correlated to fixed acidity and citric acid
and positively correlated with volatile acidity.
4. Fixed acidity is very strongly correlated to citric acid and density of the 
wine.
5. Total and free sulfur dioxide is strongly correlated with each other.

As a second step, I will make a detailed analysis of these correlations in below 
chapters.



## The relationship between pH and Quality:

```{r, echo=FALSE}

RedWineQuality$quality.factor <- as.factor(RedWineQuality$quality)

ggplot(aes(x=quality.factor, y= pH), data=RedWineQuality) +
  geom_jitter() +
  geom_boxplot()+
  stat_summary(fun.y = mean, geom = "point", shape = 18, 
               size = 2.5, color = 'red') +
  theme(text=element_text(size=10), plot.title=
          element_text(size=12, face='bold',hjust = 0.5)) +
  labs(title='pH vs Quality', x= 'Quality', y = 'pH')

```

Inspecting the boxplots for each quality level, it can be observed that pH 
value is slightly decreasing while the quality of wine increases.

Contrary to my expectation, there is a weak correlation between the pH value and
the quality. In the prior research about wine quality, it is said that pH value
is important for the quality of wine.

## The relationship between Fixed Acidity and Quality:

```{r, echo = FALSE}

ggplot(aes(x=quality.factor, y= fixed.acidity), data=RedWineQuality) +
  geom_jitter() +
  geom_boxplot()+
  stat_summary(fun.y = mean, geom = "point", shape = 18, 
               size = 2.5, color = 'red') +
  theme(text=element_text(size=10), plot.title=
          element_text(size=12, face='bold',hjust = 0.5)) +
  labs(title='Fixed Acidity vs Quality', x= 'Quality', y = 'Fixed Acidity')


```

In general positive correlation is observed between Quality and Fixed Acidity. 
Fixed acidity in the wine increases in higher quality wines. However, there is
not a continuous increase as fixed acidity values are nearly same in quality 3
and 8 wine samples. 

## The relationship between Volatile Acidity and Quality:

```{r, echo=FALSE}

ggplot(aes(x=quality.factor, y= log(volatile.acidity+1)), data=RedWineQuality) +
  geom_jitter() +
  geom_boxplot()+
  stat_summary(fun.y = mean, geom = "point", shape = 18, 
               size = 2.5, color = 'red') +
  theme(text=element_text(size=10), plot.title=
          element_text(size=12, face='bold',hjust = 0.5)) +
  labs(title='Volatile Acidity vs Quality', x= 'Quality', 
       y = 'log(Volatile Acidity')

#The strongest correlation between quality and log(volatile acidity)

```

There is very strong negative relationship between the volatile acidity and the
quality of the wine. As the quality increases, the volatile acidity is sharply
decreasing.

```{r, echo=FALSE}

ggplot(aes(x=quality.factor, y= citric.acid), data=RedWineQuality) +
  geom_jitter() +
  geom_boxplot()+
  stat_summary(fun.y = mean, geom = "point", shape = 18, 
               size = 2.5, color = 'red') +
  theme(text=element_text(size=10), plot.title=
          element_text(size=12, face='bold',hjust = 0.5)) +
  labs(title='Citric Acid vs Quality', x= 'Quality', y = 'Citric Acidity')

```

There is a positive relationship between the quality and citric acid value of
wine. 

Throughout the assessment of relationship between quality and acidity variables,
it is observed that the most strong relationship is between volatile acidity and
quality. In order to make better understandable visual analysis, I decided to 
use Volatile Acidity as the acidity variable.


## The relationship between Chloride and Quality:


```{r, echo=FALSE}

ggplot(aes(x=quality.factor, y= chlorides), data=RedWineQuality) +
  ylim(0,0.2) +
  geom_jitter() +
  geom_boxplot()+
  stat_summary(fun.y = mean, geom = "point", shape = 18, 
               size = 2.5, color = 'red') +
  theme(text=element_text(size=10), plot.title=
          element_text(size=12, face='bold',hjust = 0.5)) +
  labs(title='Chlorides vs Quality', x= 'Quality', y = 'chlorides')

#correlation between 1/chloride

```

A very weak negative relationship is found between the quality of wine and the chlorides
level in it. 

Please note that in order to display the ditribution of chlorides, y scale is 
limited to exclude far outliers.


## The relationship between Total sulfur Dioxide and Quality:

```{r, echo=FALSE}

ggplot(aes(x=quality.factor, y= total.sulfur.dioxide), data=RedWineQuality) +
  geom_jitter() +
  ylim(0,180) +
  geom_boxplot()+
  stat_summary(fun.y = mean, geom = "point", shape = 18, 
               size = 2.5, color = 'red') +
  theme(text=element_text(size=10), plot.title=
          element_text(size=12, face='bold',hjust = 0.5)) +
  labs(title='Quality vs Total Sulfur Dioxide', y= 'total sulfur dioxide',
       x = 'Quality')


```

There seems a weak negative linear relationship between total sulfur dioxide 
and quality. The total sulfur dioxide is not chaging strongly between different
quality levels.


```{r, echo=FALSE}

ggplot(aes(x=quality.factor, y= free.sulfur.dioxide), data=RedWineQuality) +
  geom_jitter() +
  ylim(0,60) +
  geom_boxplot(aes(alpha=0.05))+
  stat_summary(fun.y = mean, geom = "point", shape = 18, 
               size = 2.5, color = 'red') +
  theme(text=element_text(size=10), plot.title=
          element_text(size=12, face='bold',hjust = 0.5)) +
  labs(title='Quality vs Free Sulfur Dioxide', y= 'free sulfur dioxide',
       x = 'Quality')


```


The same situation with the total sulfur dioxide is observed also with the 
free sulfur dioxide. There is a weak corelation between the sulfur dioxide
levels and the quality.

It can also be said that the number of samples in quality levels 3,4 and 5 are
very low compared to 5,6 and 7. It would be a good idea to analyse the 
relationship between (total and free) sulfur dioxide and quality by adding more
samples with quality levels 3, 4 and 8.

Another observation from the heatmap and the last 2 plots above, total and free
sulfur dioxide levels are correlated. Note that the change pattern against 
quality are very similar.

Let's make another analysis to see total and free sulfur dioxide correlation
explicitly:

```{r, echo=FALSE}

ggplot(aes(x=free.sulfur.dioxide,y=total.sulfur.dioxide), data=RedWineQuality)+
  xlim(1, quantile(RedWineQuality$free.sulfur.dioxide, 0.95)) +
  ylim(6, quantile(RedWineQuality$total.sulfur.dioxide, 0.95)) +
  geom_jitter() +
  geom_smooth(method='lm',color='blue') +
  theme(text=element_text(size=10), plot.title=element_text(size=12, face='bold',hjust = 0.5)) +
  labs(title='Total vs Free Sulfur Dioxide', x= 'Free Sulfur Dioxide', y = 'Total Sulfur Dioxide')
  

```

It is observed that total and free sulfur dioxide levels in wine samples are
in strong positive correlation. This finding also coincides with the analysis
of quality vs total and free sulfur dioxide variables.


## The relationship between Alcohol and Quality:


```{r, echo=FALSE}

ggplot(aes(x=quality.factor, y= alcohol), data=RedWineQuality) +
  geom_jitter() +
  geom_boxplot()+
  stat_summary(fun.y = mean, geom = "point", shape = 18, 
               size = 2.5, color = 'red') +
  theme(text=element_text(size=10), plot.title=
          element_text(size=12, face='bold',hjust = 0.5)) +
  labs(title='Quality vs Alcohol', x= 'quality',
       y = 'Alcohol')

```

There is a strong relatipnship between the alcohol level and quality of the wine.
We can see lower (than expected) alcohol (mean and median) values for quality 
level 5.


## The relationship between Sulphate and Quality:

```{r, echo=FALSE}

ggplot(aes(x=quality.factor, y= sulphates), data=RedWineQuality) +
  geom_jitter() +
  geom_boxplot()+
  stat_summary(fun.y = mean, geom = "point", shape = 18, 
               size = 2.5, color = 'red') +
  theme(text=element_text(size=10), plot.title=
          element_text(size=12, face='bold',hjust = 0.5)) +
  labs(title='Quality vs Sulphate', y= 'Sulphate',
       x = 'Quality')

```


The quality of the wine is positively affected by sulphate level inside. 


## The relationship between Fixed Acidity, Volatile acidity, Citric Acid and Quality:

##1. The relationship between pH and Fixed Acidity

```{r, echo=FALSE}

ggplot(aes(x=fixed.acidity,y=pH),data=RedWineQuality) +
  geom_jitter() +
  xlim(4.6, quantile(RedWineQuality$fixed.acidity, 0.95)) +
  ylim(2.74, quantile(RedWineQuality$pH, 0.95)) +
  geom_smooth(method='lm', color = 'red') +
  theme(text=element_text(size=10), plot.title=element_text(size=12, face='bold',hjust = 0.5)) +
  labs(title='pH vs Fixed Acidity', x= 'Fixed Acidity', y = 'pH')

```

There is a strong negative correlation between pH and fixed acidity. As the 
fixed acidity increases, the pH value decreases steadily. 


##2. The relationship between pH and Volatile Acidity
 

```{r, echo=FALSE}

q1 <- ggplot(aes(x=volatile.acidity,y=pH),data=RedWineQuality) +
  geom_jitter() +
  xlim(0.12, quantile(RedWineQuality$volatile.acidity, 0.95)) +
  ylim(2.74, quantile(RedWineQuality$pH, 0.95)) +
  geom_smooth(method='lm', color = 'blue') +
  theme(text=element_text(size=10), plot.title=element_text(size=12, face='bold',hjust = 0.5)) +
  labs(title='pH vs Volatile Acidity', x= 'Volatile Acidity', y = 'pH')

q2 <- ggplot(aes(x=volatile.acidity,y=sqrt(pH)),data=RedWineQuality) +  
  geom_jitter() +
  xlim(0.12, quantile(RedWineQuality$volatile.acidity, 0.95)) +
  geom_smooth(method='lm', color = 'blue') +
  theme(text=element_text(size=10), plot.title=element_text(size=12, face='bold',hjust = 0.5)) +
  labs(title='sqrt(pH) vs Volatile Acidity', x= 'Volatile Acidity', y = 'pH')

q3 <- ggplot(aes(x=volatile.acidity,y=log10(pH)),data=RedWineQuality) +  
  geom_jitter() +
  xlim(0.12, quantile(RedWineQuality$volatile.acidity, 0.95)) +
  geom_smooth(method='lm', color = 'blue') +
  theme(text=element_text(size=10), plot.title=element_text(size=12, face='bold',hjust = 0.5)) +
  labs(title='log(pH) vs Volatile Acidity', x= 'Volatile Acidity', y = 'pH')

grid.arrange(q1,q2,q3, ncol=2)

```

pH and volatile acidity are in positive correlation. I tried to check if there is
a stronger relations exist between different transformation of pH an volatile
acidity. However, there seems no important difference between trials.

##3. The relationship between Volatile and Fixed Acidity

```{r, echo=FALSE}

ggplot(aes(x=fixed.acidity,y=volatile.acidity),data=RedWineQuality) +
  geom_jitter() +
  xlim(4.6, quantile(RedWineQuality$fixed.acidity, 0.95)) +
  ylim(0.12, quantile(RedWineQuality$volatile.acidity, 0.95)) +
  geom_smooth(method='lm', color = 'red') +
  theme(text=element_text(size=10), plot.title=element_text(size=12, face='bold',hjust = 0.5)) +
  labs(title='Volatile vs Fixed Acidity', x= 'Fixed Acidity', y = 'Volatile Acidity')

```

There is a weak negative correlation between Volatile and Fixed Acidity levels.


##4. The relationship between pH and Citric Acid

```{r, echo=FALSE}

ggplot(aes(x=citric.acid,y=pH),data=RedWineQuality) +
  xlim(0, quantile(RedWineQuality$citric.acid, 0.95)) +
  ylim(2.74, quantile(RedWineQuality$pH, 0.95)) +
  geom_jitter()+
  geom_smooth(method='lm', color = 'red') +
  theme(text=element_text(size=10), plot.title=element_text(size=12, face='bold',hjust = 0.5)) +
  labs(title='pH vs Citric Acid', x= 'Citric Acid', y = 'pH')


```

There is a negative relationship between pH value and citric acid. The pH value
decreases as the citric acid increases. 

The relationship between the 2 values can be visibly observed in the plot easily.

##5. The relationship between pH and Total Sulfur Dioxide

```{r, echo=FALSE}

ggplot(aes(x=total.sulfur.dioxide,y=pH),data=RedWineQuality) +
  xlim(6, quantile(RedWineQuality$total.sulfur.dioxide, 0.95)) +
  ylim(2.74, quantile(RedWineQuality$pH, 0.95)) +
  geom_jitter()+
  geom_smooth(method='lm', color = 'red') +
  theme(text=element_text(size=10), plot.title=element_text(size=12, face='bold',hjust = 0.5)) +
  labs(title='pH vs Total Sulfur Dioxide', x= 'Total Sulfur Dioxide', y = 'pH')
  
```

There is no relationship between the pH value and total sulfur dioxide. Since 
sulfur may be related to affect the pH of solvent in chemistry, I suspected of 
a relationship. However, pH value is independent from total sulfur dioxide.



# Bivariate Analysis


### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

The response variable quality is positively correlated with fixed acidity, 
citric acid and alcohol. In higher quality wines, alcohol an fixed acid is 
found more.

On the other hand, the quality of the wine decreases when volatile acid, 
chloride and total sulfur dioxide increases in the wine. There is also a weak
negative correlation between pH and the quality of the wine. We can say that
high quality wines have lower pH values.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

I checked the relationship between pH and other acidity variables. The analysis
shows us that pH is strongly negative correlated with fixed and citric acid 
variables, whereas it is positively correlated with volatile acidity. I also 
suspected if there is any corelation between pH and total sulfur dioxide but 
there is no such evidence.

Also I assessed the relationship between the total and free sulfur dioxide and 
there seems a strong positive relationship between these 2 variables.

### What was the strongest relationship you found?

The strongest relationship is between the alcohol and the quality of the wine.

# Multivariate Plots Section


In this chapter, I will analyze the relationships between muliple variables 
together in a plot. 


## The relationship between Quality, Acidity and Alcohol


```{r echo=FALSE, Multivariate_Plots}

RedWineQuality$alcohol.bucket <- cut(RedWineQuality$alcohol,c(8,9.7,10.8,15))
table(RedWineQuality$alcohol.bucket)

ggplot(aes(y=log10(volatile.acidity+1),x=quality.factor), data=RedWineQuality)+
  ylim(0.05,0.3) +
  geom_jitter(aes(color=alcohol.bucket, alpha=0.05))+
  geom_boxplot(aes(alpha=0.05))+
  stat_summary(fun.y = mean, geom = "point", shape = 18, 
               size = 2.5, color = 'red') +
  theme(text=element_text(size=10), plot.title=
          element_text(size=12, face='bold',hjust = 0.5)) +
  labs(title='Quality vs Acidity and Alcohol', y= 'Log Volatile Acidity', 
       x = 'Quality')
  
```

The plot above is created by applying following steps:
1. Bivariate plot of Volatile Acidity vs Quality is used as a base.
2. The continuous varible Alcohol is transformed into a categorical variable. In
order to simplify the visibility and prevent complexity, 3 categories of alcohol
are created. 
3. The points are colored according to alcohol categories.

Tt is observed that the alcohol level is increasing and (log of) volatile 
acidity is decreasing as the quality of wine increases. (Note that the color 
pattern is changing from red to blue as the quality increases.) 

As a consequence, the volatile acidity and alcohol are changing reversely as the
quality of wine changes.


## The relationship between Quality, Acidity and Total Sulfur Dioxide


```{r, echo=FALSE}

RedWineQuality$total.sulfur.dioxide.bucket <- 
  cut(RedWineQuality$total.sulfur.dioxide,c(6,26,52,289))
table(RedWineQuality$total.sulfur.dioxide.bucket)


ggplot(aes(y=log10(volatile.acidity+1),x=quality.factor), 
       data=subset(RedWineQuality, !is.na(total.sulfur.dioxide.bucket)))+
  ylim(0.05,0.3) +
  geom_jitter(aes(color=total.sulfur.dioxide.bucket, alpha=0.05))+
  geom_smooth(method='lm',color='red') +
  theme(text=element_text(size=10), plot.title=
          element_text(size=12, face='bold',hjust = 0.5)) +
  labs(title='Quality vs Acidity and Total Sulfur Dioxide', 
       y= 'Log Volatile Acidity', x = 'Quality')


```

In order to create the above plot, the following steps are followed:
1. Bivariate plot of Volatile Acidity vs Quality is used as a base.
2. The continuous varible Total sulfur dioxide is transformed into a categorical
variable. In order to simplify the visibility and prevent complexity, 3 
categories of total sulfur dioxide are created. 
3. The points are colored according to total sulfur dioxide categories.

In the plot above, the patterns of change of total sulfur dioxide and volatile
acidity according to the quality is analysed.

It is observed that the blue color is becoming rare in small amount as the 
quality level increases, which is an expected result of weak negative 
correlation between the qualit and total sulfur dioxide. 

However, there is no change in color pattern in vertical axis. It means that 
there is no correlation between sulfur dioxide and volatile acidity. This result
also coincides with the heat map.


## The relationship between Quality, Acidity and Chlorides

```{r, echo=FALSE}

RedWineQuality$chlorides.bucket <- 
  cut(RedWineQuality$chlorides,c(0.01,0.073,0.085, 0.62))
table(RedWineQuality$chlorides.bucket)


ggplot(aes(y=log10(volatile.acidity+1),x=quality.factor), 
       data=subset(RedWineQuality, !is.na(chlorides.bucket)))+
  geom_jitter(aes(color=chlorides.bucket, alpha=0.05))+
  geom_smooth(method='lm',color='red') +
  theme(text=element_text(size=10), plot.title=
          element_text(size=12, face='bold',hjust = 0.5)) +
  labs(title='Quality vs Acidity and Chlorides', y= 'Log Volatile Acidity', 
       x = 'Quality')



```

In order to create the above plot, the following steps are followed:
1. Bivariate plot of Volatile Acidity vs Quality is used as a base.
2. The continuous varible Chlorides is transformed into a categorical
variable. In order to simplify the visibility and prevent complexity, 3 
categories of total sulfur dioxide are created. 
3. The points are colored according to Chlorides categories.

There is no visually observable color change in y and x axes. We can conclude 
that there is not a strong relationship between chlorides and other 2 variables
quality and volatile acidity. Chlorides level does not change meaningfully with 
as quality or volatile acidity changes.


## Linear model depicting the relationship between Quality, Volatile Acidity, Alcohol, Total Sulfur Dioxide and Chlorides

A linear model is created to formulate the linear relationship between Quality 
and the explanatory variables of Volatile Acidity, Alcohol, Total Sulfur Dioxide
and Chlorides.


```{r, echo=FALSE}

m1 <- lm(I(quality)~I(log(volatile.acidity)), data=RedWineQuality)
m2 <- update(m1, ~.+alcohol)
m3 <- update(m2, ~.+total.sulfur.dioxide)
m4 <- update(m3, ~.+chlorides)
mtable(m1,m2,m3,m4)

```


quality = 2.207 + -0.671xlog(volatile acidity+1) + 0.295xalcohol + -0.002x
total sulfur dioxide + -0.305xchlorides

However, when we check the r-squared values, it seems that the linear model is
far from covering the variance. This model can not be used to predict the 
quality level of wine with input values of Volatile Acidity, Alcohol, Total 
Sulfur Dioxide and Chlorides.


## The relationship between pH, Fixed Acidity, Volatile Acidity and Citric Acid

```{r, echo=FALSE}

#pH, volatile.acidity, fixed.acidity, citric.acid

RedWineQuality$fixed.acidity.bucket <- 
  cut(RedWineQuality$fixed.acidity, c(4.5, 7.10,8.32,16))
summary(RedWineQuality$fixed.acidity.bucket)

RedWineQuality$citric.acid.bucket <- 
  cut(RedWineQuality$citric.acid, c(-1, 0.09,0.26,0.42,1))
summary(RedWineQuality$citric.acid.bucket)

ggplot(aes(x=log10(volatile.acidity+1), y= pH), data=RedWineQuality)+
  ylim(3, 3.5) +
  xlim(0.1, 0.3) +
  geom_jitter(aes(shape=factor(fixed.acidity.bucket),color=citric.acid.bucket)) +
  geom_smooth(method='lm',color='blue') +
  theme(text=element_text(size=10), plot.title=
          element_text(size=12, face='bold',hjust = 0.5)) +
  labs(title='pH vs Volatile Acidity, Fixed Acidity and Citric Acid', 
       x= 'Log Volatile Acidity', y = 'pH')

```

In order to create the above plot, the following steps are followed:
1. Bivariate plot of pH and Log of volatile acidity are used as a base.
2. The continuous varible Fixed acidity and citric acid are transformed into a 
categorical variables. In each category 3 levels are set.
3. The points are colored according to citric acid categories.
4. The shapes of posint are differentiated according to fixed acid categories.

Visual observations depict these findings:
1. pH is in positive correlation with volatile acidity.

2. Citric acid is in negative correlation with both pH and volatile acidity. 
Note that the color is changing in both x and y axes:
   i. Color is transforming from blue and viloet to red and green from left to 
   right, which means a downward change in citric acid level. 
   ii. Color is becoming blue and red from viloet and green from down to top, 
   which also means a decreasing citric acid level as pH increases. 

3. Fixed acidity is in negative correlation with both pH and volatile acidity.
  i. Note that there are mostly square shapes in the bottom left of the plot.
  This is where the volatile acidity and pH are both in low levels.
  ii. The square levels disappear as pH or volatile acidity increases in y 
  or x axis. There are more triangles in bottom right and upper left corners.
  iii. The fixed acidity is lowest in the upper right corner where the pH and
  volatile acidity are both highest.


4. Fixed acidity is in positive correlation with citric acid. 

## Linear model depicting the relationship between pH, Volatile Acidity, Fixed Acidity and Citric Acid

A linear model is built to estimate the pH value with 3 other acidity variables 
given as inputs.

```{r, echo=FALSE}

m1 <- lm(I(pH)~I(log(volatile.acidity+1)), data=RedWineQuality)
m2 <- update(m1, ~.+fixed.acidity)
m3 <- update(m2, ~.+citric.acid)
mtable(m1,m2,m3)

```

pH = 3.764 + -0.01xlog(volatile acidity+1) + -0051xfixed acidity + 
-0.124xcitric acid.

The r-squared value is 0.479 which is much less than 1 which means that the model
is far from estimating the pH value with 3 acidity variables as input.


# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

In the "Quality vs Acidity vs Alcohol" plot, I have sketched the change of 
alcohol value in the samples on the original "Quality vs Volatile Acidity"
plot. It is displayed that alcohol value does not change with volatile 
acidity but the color only changes as the quality varies.

A similar relationship is displayed in the following 2 plots of displaying 
the relationship between quality, volatile acidity, total sulfur dioxide and 
chlorides. The distributions of Both total sulfur dioxide and chlorides 
does not change with volatile acidity but it decreases only with quality. 

The last plot is displaying the relationship between pH, volatile acidity,
fixed acidity and citric acid:

1. There is negative correlation between pH and volatile acidity.
2. Citric acid decreases with increasing pH and volatile acidity: Blue color
is disappearing and turning into red.
3. Fixed acidity decreases as pH and volatile acidity increases: The square 
shape is transforming into circles and triangles which resembe lower values of 
fixed acid.

### Were there any interesting or surprising interactions between features?

All of the findings are coinciding with the findings in bivariate analysis.

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

I created 2 linear models:

1. A linear model with inputs log of volatile acidity, alcohol, total sulfur
dioxide and chlorides, and quality as output.

My linear models has some drawbacks:

- I only took some of the variables included in the dataset to construct the 
model. For example, I could add the density vaiable also. 

- Other factors which definitely affect the quality of a wine are not included in
tha dataset. The wineyard, kind of grape, different flavors, bottling process
may also be important.

- If all other factors are included, the reliability of the model would increase.

2. A linear model with fixed and volatile acidity and citric acid as 
inputs and pH as output.

- The second model may also require extra variables to be reliable since the 
r-squared values seem low. 

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}

q1 <- qplot(x=total.sulfur.dioxide, data=RedWineQuality, , binwidth = 3) +
  scale_x_continuous(limits=c(6,150)) +
  theme(text=element_text(size=10), plot.title=
          element_text(size=12, face='bold',hjust = 0.5)) +
  labs(title='Total Sulfur Dioxide Histogram', x= 'Total Sulfur Dioxide', 
       y = 'Count')

q2 <- ggplot(aes(x=total.sulfur.dioxide,y=quality),data=RedWineQuality) +
  xlim(0,150) +
  geom_jitter() +
  geom_smooth(method='lm', color = 'red') +
  theme(text=element_text(size=10), plot.title=
          element_text(size=12, face='bold',hjust = 0.5)) +
  labs(title='Quality vs Total Sulfur Dioxide', x= 'total sulfur dioxide', 
       y = 'Quality')

grid.arrange(q1,q2)

```

### Description One

It is an interesting finding that total and free sulfur dioxide which are in
negative correlation with the quality of the wine have right skewed histograms, 
in contrat with other variables such as acidity, chlorides and alcohol. 

It means that most of the wine samples have low sulfur dioxide levels and this 
may not be used as a very strong differentiator for setting quality of wine.

This finding is interesting to see the effect of a variable with right or left 
skewed histogram on the resultant variable with a normal distribution. 


### Plot Two
```{r echo=FALSE, Plot_Two}

RedWineQuality$alcohol.bucket <- cut(RedWineQuality$alcohol,c(8,9.7,10.8,15))
table(RedWineQuality$alcohol.bucket)

ggplot(aes(x=log10(volatile.acidity+1),y=quality), data=RedWineQuality)+
  xlim(0.05,0.3) +
  geom_jitter(aes(color=alcohol.bucket, alpha=0.05))+
  geom_smooth(method='lm',color='red') +
  theme(text=element_text(size=10), plot.title=
          element_text(size=12, face='bold',hjust = 0.5)) +
  labs(title='Quality vs Acidity and Alcohol', x= 'Log Volatile Acidity', 
       y = 'Quality')
  

```

### Description Two

The plot "Quality vs Acidity and Alcohol" provided me expected results. The
quality increases with alcohol and decreases with acidity.

However, acidity and alcohol are not in a direct relationship. This is expected
as alcohol and acidity are independent groups of molecules in chemistry where 
most people think that acidic liquids contains alcohol as people think them 
bitter.

### Plot Three
```{r echo=FALSE, Plot_Three}

RedWineQuality$fixed.acidity.bucket <- 
  cut(RedWineQuality$fixed.acidity, c(4.5, 7.10,8.32,16))
summary(RedWineQuality$fixed.acidity.bucket)

RedWineQuality$citric.acid.bucket <- 
  cut(RedWineQuality$citric.acid, c(-1, 0.09,0.26,0.42,1))
summary(RedWineQuality$citric.acid.bucket)

ggplot(aes(x=log10(volatile.acidity+1), y= pH), data=RedWineQuality)+
  ylim(3, 3.5) +
  xlim(0.1, 0.3) +
  geom_jitter(aes(shape=factor(fixed.acidity.bucket),color=citric.acid.bucket)) +
  geom_smooth(method='lm',color='blue') +
  theme(text=element_text(size=10), plot.title=
          element_text(size=12, face='bold',hjust = 0.5)) +
  labs(title='pH vs Volatile Acidity, Fixed Acidity and Citric Acid', 
       x= 'Log Volatile Acidity', y = 'pH')


```

### Description Three

This plot explains the relationship between different variables and the pH value
of the wine. pH value is positively correlated with volatile and citric acid but
in negative correlation with fixed acidity. Also, I observed the negative 
correlation between the fixed and volatile acidity. 

This plot provided me to see the effects of 3 explanatory variables on pH as 
the resultant variable.

However, the quality of wine is not strongly correlated with pH value but in 
stronger correlation with volatile acidity. Therefore. I used volatile acidity
to emphasize the relationship between acidity and quality.


------

# Reflection

> **Tip**: Here's the final step! Reflect on the exploration you performed and
the insights you found. What were some of the struggles that you went through?
What went well? What was surprising? Make sure you include an insight into
future work that could be done with the dataset.
